---
title: "Joint Spatiotemporal Model"
subtitle: "Hierarchical Bayesian Inference for NWS Detection and Abundance"
date: last-modified
format:
  html: 
    df-print: kable
    code-fold: show
    code-summary: "Show R Code"
    code-overflow: wrap
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    html-math-method: katex
    theme: cosmo
---

```{=html}
<style type="text/css">
  body { font-size: 11pt; }
  h1, h2, h3 { color: #2c3e50; }
  .math { font-size: 1.1em; }
  code.r { font-size: 9pt; }
</style>
```

## Overview
This script demonstrates the **Bayesian hierarchical joint model** used to estimate New World Screwworm (NWS) intensity. Given that detections arise from opportunistic surveillance, the model explicitly separates the **observation process** (detection probability) from the **biological process** (abundance) to isolate the true biological signal from surveillance bias.

---

## Computational & Privacy Note
This script is for **demonstrative purposes only** and has not been executed within this document. The full joint model is computationally intensive, containing approximately 10.2 million latent entries and 260,185 replicated spatial field elements. Due to these requirements and the sensitive nature of the original surveillance coordinates, which involve private residences and businesses, the code provided here tracks the structural methodology but does not include the raw datasets.

## HPC Execution Environment
To handle the high dimensionality of the latent Gaussian fields and the spatial point-process likelihood, the model was executed on a high-performance computing (HPC) cluster. The following SLURM batch script illustrates the resource allocation required for a successful run once reasonable starting values (*theta_init* below) have been identified:

```bash
#!/bin/bash
#SBATCH --job-name=nws_joint_model
#SBATCH -A disease_ecology
#SBATCH -N 1
#SBATCH -n 1
#SBATCH -p bigmem
#SBATCH --cpus-per-task=14
#SBATCH -t 35:59:00
#SBATCH --mem=280G

# Load necessary spatial and computational modules
module load udunits proj geos/3.12.1 gdal/3.8.5 openblas r/4.4

# Set threading for INLA/OpenBLAS to match allocated CPUs
export OMP_NUM_THREADS=14
export OPENBLAS_NUM_THREADS=14

# Execute the R script in batch mode
R CMD BATCH --no-save --no-restore jcat.R
```

---

## Statistical Formulation
The model evaluates two coupled processes across 15,305 spatial mesh nodes ($i$) over epidemiological weeks ($t$):

$$
\begin{aligned}
y_{1,it} &\sim \text{Bernoulli}(p_{it}) && \text{(Detection Tier)} \\
y_{2,it} &\sim \text{NegBinom}(\mu_{it}, n) && \text{(Abundance Tier)}
\end{aligned}
$$

### Linear Predictors
The expected outcomes are linked to predictors via logit and log link functions:

**Detection Tier (Observation):**
$$\text{logit}(p_{it}) = \beta_{0}^{(1)} + \sum \beta_{1} \mathbf{X}_{1,it} + \xi_{1,it} + w_{1,t} + \gamma_{c(i)}$$

**Abundance Tier (Biological):**
$$\log(\mu_{it}) = \log(\text{Area}_i) + \beta_{0}^{(2)} + \sum \beta_{2} \mathbf{X}_{2,it} + f(\text{H}_{it}) + \xi_{2,it} + \alpha \xi_{1,it} + w_{2,t}$$  
**Key Components:**  
* $\mathbf{X}_{1}, \mathbf{X}_{2}$: Linear covariates (e.g., environment, infrastructure).  
* $\xi_{1,it}, \xi_{2,it}$: Replicated spatial Gaussian Fields (SPDE).  
* $\alpha \xi_{1,it}$: A shared spatial component where $\alpha$ scales quantifies detection process on abundance.  
* $w_{1,t}, w_{2,t}$: Temporal random walks (RW1) for weekly trends.  
* $\gamma_{c(i)}$: IID random effect for administrative jurisdictions.  
* $f(\text{H}_{it})$: Non-linear livestock density effect (RW2).  
  

> **Note on $\alpha \xi_{1,it}$:** This represents a shared spatial component where $\alpha$ quantifies the association between surveillance effort and biological intensity, allowing for the correction of preferential sampling.

---

## Spatial Projection
We utilize the **Stochastic Partial Differential Equation (SPDE)** approach with Penalized Complexity (PC) priors to prevent overfitting.

```{r eval=FALSE}
# SPDE for Tier 1: Detection (xi_1)
spde.t1 <- inla.spde2.pcmatern(mesh.dom, 
                               alpha = 2,
                               prior.range = c(90, 0.5),  
                               prior.sigma = c(1, 0.5),
                               constr = TRUE)

# SPDE for Tier 2: Abundance (xi_2)
spde.t2 <- inla.spde2.pcmatern(mesh.dom, 
                               alpha = 2,
                               prior.range = c(150, 0.05),  
                               prior.sigma = c(0.5, 0.05),
                               constr = TRUE)


# Temporal steps
t_steps <- length(unique(tier_1.set$time_index))

# Define Spatial Indices
field.xi1 <- inla.spde.make.index("field.xi1", n.spde = spde.t1$n.spde, n.group = t_steps)
field.xi2 <- inla.spde.make.index("field.xi2", n.spde = spde.t2$n.spde, n.group = t_steps)
```

## Data Integration

### Detection Stack
```{r eval=FALSE}
A.xi1 <- inla.spde.make.A(mesh.dom, loc = locs_t1, group = tier_1.set$time_index)

eff_t1 <- list(
  c(field.xi1, list(intercept1 = 1)), 
  list(north = tier_1.set$northing_km_s,
       road_dens = tier_1.set$road_dens_s,
       night_illum = tier_1.set$night_illum_s,
       admin_gamma = as.integer(as.factor(tier_1.set$admin_u)),
       week_w1 = tier_1.set$timestep)
)

jtier1.stk <- inla.stack(
  data = list(Y = cbind(tier_1.set$Yi_cens_test, NA), link = 1, e = rep(NA, nrow(tier_1.set))),
  A = list(A.xi1, 1), 
  effects = eff_t1,   
  tag = "det.stack"
)
```

### Abundance Stack
```{r eval=FALSE}
A.xi2 <- inla.spde.make.A(mesh.dom, loc = locs_t2, group = tier_2.set$time_index)

eff_t2 <- list(
  c(field.xi2, list(field.xi1.copy = field.xi1$field.xi1), list(intercept2 = 1)), 
  list(mintemp = tier_2.set$mintemp_s,
       soilmoist = tier_2.set$soilmoist_s,
       leafarea = tier_2.set$leafarea_s,
       livestock_H = tier_2.set$cattle20_q,
       week_w2 = tier_2.set$timestep)
)

jtier2.stk <- inla.stack(
  data = list(Y = cbind(NA, tier_2.set$Count_cens_test), link = 2, e = tier_2.set$area),
  A = list(A.xi2, 1), 
  effects = eff_t2,   
  tag = "abun.stack"
)
```

### Join Data  
The tier specific stacks are merged into a single object for the *inla()* call. This allows the model to estimate both processes simultaneously, propagating uncertainty from the detection tier to the abundance tier.
```{r eval=FALSE}

joint_stk <- inla.stack(jtier1.stk, jtier2.stk)
```

## Model Fitting
The model is fitted using a joint likelihood (Bernoulli and Negative Binomial). We utilize the *eb* (Empirical Bayes) integration strategy to handle the high dimensionality of the latent fields.  
  

### Priors
Penalized Complexity (PC) priors are defined for the random effects to prevent overfitting and reflect known dispersal capacities.

```{r eval=FALSE}
# Shared spatial field scaling (alpha) and fixed effects
hyper_alpha <- list(beta = list(prior = "normal", param = c(0, 1)))

# PC priors for Random Walks (Temporal and Jurisdictional)
pc_rw <- list(prec = list(prior = "pc.prec", param = c(1, 0.01)))
pc_rw_strong <- list(prec = list(prior = "pc.prec", param = c(0.3, 0.01)))
pc_rw_cat <- list(prec = list(prior = "pc.prec", param = c(0.5, 0.05)))
```


### Formula  
The formula  couples the two tiers, using `f(..., copy = "field.xi1")` to represent the shared spatial effect $\alpha \xi_{1,it}$. 
```{r eval=FALSE}
formula_nws <- Y ~ -1 + intercept1 + intercept2 + # response and intercepts for each tier
  # Tier 1 Components
  f(field.xi1, # tier 1 field
    model = spde.t1, 
    group = field.xi1.group, 
    control.group = list(model = "iid", 
                         hyper = pc_rw)) +
  f(week_w1, 
    model = "rw1", 
    constr = TRUE, 
    scale.model = TRUE, 
    hyper = pc_rw) +
  f(admin_gamma, 
    model = "iid", 
    constr = TRUE, 
    hyper = pc_rw_strong) +
  north + road_dens + night_illum + 
  
  # Tier 2 Components
  f(field.xi2, 
    model = spde.t2, 
    group = field.xi2.group, 
    control.group = list(model = "iid", 
                         hyper = pc_rw)) +
  f(field.xi1.copy, 
    copy = "field.xi1", 
    group = field.xi1.copy.group, 
    fixed = FALSE, 
    hyper = hyper_alpha) +
  f(week_w2, model = "rw1", 
    constr = TRUE, 
    scale.model = TRUE, 
    hyper = pc_rw) +
  f(livestock_H, 
    model = "rw2", 
    constr = TRUE, 
    scale.model = TRUE, 
    hyper = pc_rw_cat) +
  mintemp + soilmoist + leafarea + rhum + cattle_s + horses_s + pigs_s + goats_s + sheep_s
```

### Run Model
```{r eval=FALSE}
# Quantiles for posterior summaries
return_quants <- c(0.025, 0.25, 0.5, 0.75, 0.975) 

# Initial values for hyperparameters (speeds up run)
theta_init <- c(4.65, 0.98, -2.01, -3.61, 2.30, 5.02, -0.02, 0.50, 5.42, 0.39)

model_joint <- inla(formula_nws,# formula
                family = c("binomial", "nbinomial"), # family for each tier
                data = inla.stack.data(joint_stk), # combined data
                E = inla.stack.data(joint_stk)$e, # geographic exposure
                control.predictor = list(A = inla.stack.A(joint_stk), 
                                         compute = TRUE, #return fitted values
                                         link = 1), 
                control.inla = list(strategy = "adaptive", 
                                    int.strategy = "eb"), # Empirical Bayes strategy
                control.mode = list(restart = TRUE, theta = theta_init), # initial values
                num.threads = 12, # run using 16 cores, others reserved for background processes
                quantiles = return_quants,
                verbose = TRUE)
```
