---
title: "Response"
subtitle: "Construct Bivariate Response Variable"
date: last-modified
format:
  html: 
    df-print: kable
    code-fold: show
    code-summary: "Show R Code"
    code-overflow: wrap
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    html-math-method: katex
    theme: cosmo
---
```{=html}
<style type="text/css">

body, td {
   font-size: 11pt;
}
code.r{
  font-size: 9pt;
}
pre {
  font-size: 11pt
}
</style>
```


## Overview
The script demonstrates how to expand and index reported NWS case detections and mesh nodes to ensure that all weeks are represented in the data set.  Additionally, the script shows how to aggregate individual detection counts to natural natural neighborhoods as needed for the abundance tier in the joint model described in the associated paper. 

Once data has been formatted correctly, it can be organized as a bivariate response variable such that **Tier 1** describes point-level NWS case detections:  
$$
y_{1it} =
\begin{cases}
1, & \text{NWS detection reported at location } i \text{ and time } t \\
0, & \text{NWS not detected or reported at location } i \text{ and time } t
\end{cases}
$$
  
and, **Tier 2** represents the number of NWS Cases reported in each neighborhood:   
$$
y_{2it} =
\begin{cases}
NA, & \text{NWS detections not reported in neighborhood } i \text{ and time } t \\
\text{Count}, & \text{Sum of all NWS detections in neighboorhood } i \text{ and time } t
\end{cases}
$$
  
  
   
   
   
           
::: {.callout-tip icon=false}   
#### Note
This script assumes that supporting data has already been downloaded to a directory called `/demo`, that NWS case detections have been simulated as described on the [Simulation tab](https://geoepi.github.io/hominivorax-geostat/simulation.html), and that the triangulated mesh and areal tessellation presented on the [tessellation tab](https://geoepi.github.io/hominivorax-geostat/spatial_domain.html) have been constructed and saved. 
:::
  
---
   
## Libraries  
Load needed libraries and packages.  
```{r, warning=FALSE, message=FALSE}
library(tidyverse) # data wrangling
options(dplyr.summarise.inform = FALSE)
library(here) # directory management
library(patchwork)# combine ggplots
library(pals) # color palettes
library(osfr) # interact with Open Science Framework (OSF, data repository)
library(terra) # spatial data manipulation
library(sf) # spatial data manipulation
library(spatstat) # spatial data manipulation
library(raster) # spatial data manipulation
select <- dplyr::select # deconflict dplyr w/raster  

# see https://www.r-inla.org/home
# install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
library(INLA) # mesh construction, inference
```

## Custom Functions 
Load customized functions.  
```{r, warning=FALSE, message=FALSE}
source(here("R/utilities.R"))
source_dir(here("R"))
```

## Load Preprocessed Data
```{r warning=FALSE, message=FALSE}
# simulated case detections
sim_data_df <- read_csv(here("demo/sim_locations_demo.csv"))

# triangulated mesh
mesh.dom <- readRDS(here("demo/mesh.rds"))

# tessellation
mesh_polys <- st_read(here("demo/neighboorhoods.gpkg"))

# study area administrative boundaries
states_proj <- st_read(here("demo/study_polygons.gpkg"))
```
  
  
Geographic projection of simulated NWS detections (currently unprojected)
```{r warning=FALSE, message=FALSE}
sim_data_pnts <- vect(sim_data_df, geom = c("x", "y"), "EPSG:4326") # to points
sim_data_pnts <- project(sim_data_pnts, states_proj) # match other data projection

# back to dataframe
sim_data_df <- as.data.frame(sim_data_pnts, geom = "xy")


```

   
## Prepare Node Locations  
```{r warning=FALSE, message=FALSE}
# get node coordinates
int_pnts <- as.data.frame(cbind(mesh.dom$loc[,1],
                                mesh.dom$loc[,2]))

names(int_pnts) <- c("x", "y")

# convert to points
int_pnts <- vect(int_pnts, geom = c("x", "y"), crs = states_proj)
```

Identify and remove nodes located in oceans (non-NWS habitat)
```{r warning=FALSE, message=FALSE}
states_simple <- vect(states_proj)
states_simple$inside <- 1
int_pnts$inside <- extract(states_simple["inside"], int_pnts)$inside

# back to dataframe
integ_df <- as.data.frame(int_pnts, geom="xy") %>%
  filter(!is.na(inside)) %>% # remove ocean nodes
  mutate(Y1 = 0, # no detection
         Exp = NA) %>% # no exposure (point locations, no area)
  select(Y1, Exp, x, y)
```


## Time Expansion
Create dataframe with all years and weeks for the study period, weeks between `2022-01-01` and `2026-03-01`.  
```{r}
study_start <- as_date("2022-01-01")
study_end <- as_date("2026-03-01")

data_weeks <- sim_data_df %>%
  distinct(year, week)

# date sequence to fill gaps with no NWS
calendar_weeks <- data.frame(date = seq(study_start, study_end, by = "day")) %>%
  mutate(
    year = epiyear(date),
    week = epiweek(date)
  ) %>%
  distinct(year, week)

time_index <- bind_rows(data_weeks, calendar_weeks) %>%
  distinct(year, week) %>%
  mutate(year = as.numeric(year), week = as.numeric(week)) %>%
  filter(year >= 2022) %>%
  arrange(year, week)

```

Replicate nodes to ensure a copy for each week in the study period.  
```{r}
quad_expand <- crossing(integ_df, time_index)

dim(quad_expand) # dimensions

head(quad_expand)
tail(quad_expand)
```

## Clean Detections
```{r}
nws_detections_df <- sim_data_df %>%
  mutate(Y1 = 1, # NWS detection was reported
         Exp = NA) %>% # no exposure (point locations, no area)
  select(Y1, Exp, x, y, year, week)
```

## Combine Detection Tier
Combine mesh nodes (quadrature) with NWS detections (simulated)
```{r}
tier1_df <- rbind(nws_detections_df, quad_expand)

# check
sum(tier1_df$Y1) == nrow(nws_detections_df)

dim(tier1_df)
```


## Aggregate Counts
Summing NWS detections within each neighborhood.
```{r}
# assign unique polygon ID
mesh_polys$poly_id <- 1:nrow(mesh_polys)

# get the polygon ID for each NWS reported detection (or node)
tier1_pnts <- vect(tier1_df, geom = c("x", "y"), crs = states_proj) # to spatial points
mesh_polys <- vect(mesh_polys) # convert to SpatVector
extracted <- extract(mesh_polys, tier1_pnts) 

# add ID to dataframe
tier1_df$poly_id <- extracted$poly_id

# count points in each polygon by year-week
counts <- tier1_df %>%
  filter(Y1 == 1) %>% # only count detections, not nodes (Y1==0)
  count(poly_id, year, week, name = "n_points")

range(counts$n_points) # max simulated detections in neighborhood and week
```

## Match to Neighborhood
Lastly, aggregate counts per neighborhood are match to the location of that neighborhood centroid.  
```{r warning=FALSE, message=FALSE}
# get centroid coordinates
centroids <- centroids(mesh_polys)
mesh_centroids_df <- as.data.frame(geom(centroids)[, c("x", "y")])
mesh_centroids_df$poly_id <- mesh_polys$poly_id

# all combinations of poly_id and time
# making a copy for each week in the study period
all_combos <- mesh_centroids_df %>%
  select(poly_id, x, y) %>%
  crossing(time_index)


# match with aggregate counts
# fill with NA if no NWS at location of time step
weekly_counts <- all_combos %>%
  left_join(counts, by = c("poly_id", "year", "week")) %>%
  arrange(poly_id, year, week)


sum(weekly_counts$n_points, na.rm=TRUE) == sum(counts$n_points) #verify all counts were matched
```

## Map Response
Viewing the detection and count data for a select year and week.  
```{r warning=FALSE, message=FALSE, fig.width=20, fig.height=10}
# convert to SF for plotting
tier1_sf <- st_as_sf(tier1_df, coords = c("x", "y"), crs = st_crs(states_proj))

tier2_sf <- mesh_polys %>% 
  st_as_sf() %>% 
  inner_join(weekly_counts, by = "poly_id")
```
```{r warning=FALSE, message=FALSE, fig.width=20, fig.height=10}
#| label: fig-resp
#| fig-cap: "Comparison of data used as bivariate response variable for detection (left) and potential abundance (right) tiers for arbitrary week. Panel at left shows point-level location of simulated detections whereas that at right shows sum of detections per neighborhood."


plot_bivariate_response(
  tier1_sf = tier1_sf,
  tier2_sf = tier2_sf,
  states_sf = states_proj,
  yr = 2025,
  wk = 35
)
```



::: {.callout-tip icon=false} 
#### Next Step
Proceed to the next step to view the [model](https://geoepi.github.io/hominivorax-geostat/model.html) specification. 
:::








