---
title: "Simulation"
description: "Source data download and simulation of NWS point detections"
format:
  html: 
    df-print: kable
    code-fold: show
    code-summary: "Hide code"
    code-overflow: wrap
    toc-title: Page Contents
    toc: true
    toc-depth: 2
    toc-location: right
    number-sections: false
    html-math-method: katex
    smooth-scroll: true
editor: source
editor_options: 
  chunk_output_type: console
---
```{=html}
<style type="text/css">

body, td {
   font-size: 11pt;
}
code.r{
  font-size: 9pt;
}
pre {
  font-size: 11pt
}
</style>
```


## Overview
The script demonstrates how to download model results archived on the Open Science Framework (OSF) and simulate New World screwworm (NWS) case detections.   
    
The downloaded data includes spatial GeoTiff files depicting potential abundance and estimated spread for each week between January 2022 and February 2026, spatial boundaries for analysis and visualization, and a copy of previously simulated NWS case point locations.   

## Libraries  
Load needed libraries and packages.  
```{r, warning=FALSE, message=FALSE}
library(tidyverse) # data wrangling
options(dplyr.summarise.inform = FALSE)
library(here) # directory management
library(pals) # color palettes
library(osfr) # interact with Open Science Framework (OSF, data repository)
library(terra) # spatial data manipulation
library(sf) # spatial data manipulation
library(spatstat) # spatial data manipulation
```

## Custom Functions 
Load customized functions.  
```{r, warning=FALSE, message=FALSE}
source(here("R/utilities.R"))
source_dir(here("R"))
```

## Download OSF Data  
Spatial GeoTIFF files (.tif) depicting NWS estimated niche capacity and spread are available on the [Open Science Framework](https://osf.io/uvqmw/).  Here we link the location and download the files to a demonstration directory (`/demo`).  
  
**Downloaded Files:**    
- *spread_GeoTIFF_2026-03-24.zip*: Zipped folder containing rasters showing estimated NWS spread.       
- *niche_GeoTIFF_2026-03-24.zip*: Zipped folder containing rasters showing potential NWS abundance.   
- *study_polygons.gpkg*: Geopackage containing *sf* class multipolygons delineating political boundaries.    
- *sim_locations_2026-03-24.csv*: Simulated NWS detections; the same file is reproduced in this script.   
   

```{r, warning=FALSE, message=FALSE}
# file location on OSF
osf_project_location <- osf_retrieve_node("https://osf.io/uvqmw/")

# Project name on OSF
osf_project_location$name 

# create demo directory
if (!dir.exists("demo")) dir.create("demo", recursive = TRUE)

# get data file names
osf_id <- osf_project_location %>%
  osf_ls_files()

# download all data
osf_download(osf_id,
             path = here("demo"),
             conflicts = "overwrite")
```

#### Unzip Grids
Spatial grids are bundled in zip files. Unzip the files.  
```{r warning=FALSE, message=FALSE}
osf_id$name # file names

unzip(file.path(here("demo"),"spread_GeoTIFF_2026-03-24.zip"), exdir = here("demo")) # spread grids
unzip(file.path(here("demo"), "niche_GeoTIFF_2026-03-24.zip"), exdir = here("demo")) # niche grids
```

#### Spatial Polygons
Read polygons defining jurisdictional and administrative boundaries
```{r, warning=FALSE, message=FALSE}
states_proj <- st_read(here("demo/study_polygons.gpkg"))
```


## Inspect Grids
Spatial grid file names are coded with the year and week they represent.  The *stack_rasters()* function will read the files from the corresponding directory, order them by year and week, and return the data as SpatRaster objects. Values in both objects provide the number of NWS detections per 25x25~km grid cell (approx. 625~km$^2$).  The *niche* product is the number of *potential* detections based on environmental conditions, whereas the *spread* product is an estimate of how many occurred between January 2022 and February 2026.    
   
Organize and stack rasters.  
```{r, warning=FALSE, message=FALSE}
niche_stk <- stack_rasters(here("demo/niche")) # reflecting niche conditions, limits, or capacity
spread_stk <- stack_rasters(here("demo/spread")) # reflecting invasion/spread
```


#### Example Niche Maps
```{r fig.width=15, fig.height=15, warning=FALSE, message=FALSE}
quick_map_plot(grid_stk=niche_stk, target_year=2024, target_week=10, boundary_polys=states_proj)
quick_map_plot(grid_stk=niche_stk, target_year=2025, target_week=30, boundary_polys=states_proj)
```

#### Example Spread Maps
```{r fig.width=15, fig.height=15, warning=FALSE, message=FALSE}
quick_map_plot(grid_stk=spread_stk, target_year=2024, target_week=10, boundary_polys=states_proj)
quick_map_plot(grid_stk=spread_stk, target_year=2025, target_week=30, boundary_polys=states_proj)
```


## Estimated Detections
The *sum_intensity_grid()* function, when applied to the estimated spread grids, will sum the estimated number of detections for each week.  This approximates the mean number of cases as predicted by the model described in the manuscript. Note that the function includes an input option `min_est=` that specifies the minimum cell value to consider a detection.  
   
The dataframe returned by the function includes a column *count* that will be used to simulate spatial points in the next step.   
```{r fig_width=15, fig.height=10, warning=FALSE, message=FALSE}
weekly_counts <- sum_intensity_grid(spread_stk, min_est = 1)

ggplot(data = weekly_counts, aes(x = date, y = count)) +
  geom_col(fill = "steelblue", alpha = 0.6, width = 7) + 
  theme_minimal() +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") + 
  scale_y_continuous(breaks=scales::pretty_breaks(6)) + 
  theme(
    plot.margin = unit(c(5, 0.01, 5, 0.01), "cm"),
    legend.position = " none",
    legend.background = element_rect(fill = "white", color = "grey90"),
    plot.title = element_text(face = "bold", size = 16),
    axis.title   = element_text(size = 22, face = "bold"),
    axis.text.x    = element_text(size = 16, face = "bold", angle = 45, hjust = 1),
    axis.text.y    = element_text(size = 16, face = "bold")
  ) +
  labs(
    title = " ",
    subtitle = " ",
    x = " ",
    y = "Estimated Case Count",
    caption = " "
  )

```


## Simulate Spatial Points
The *simulate_detection_points()* function will use the [**spatstat** package](https://spatstat.org/) to generate random spatial points based on target weekly counts given in the *weekly_counts* dataframe created in the prior step, and estimated intensity as given in the spread grids (*spread_stk*).  Values in *spread_stk* serve as weights for probabilistic assignment of point locations.       
  
The returned result is a dataframe with x and y coordinates, year, and week.  Note that the projected coordinate system is used by default.  Conversion to latitude and longitude can be performed as shown below. 
```{r, warning=FALSE, message=FALSE}
sim_points <- simulate_detection_points(raster_stack=spread_stk,
                                        obs_df=weekly_counts, 
                                        rseed=1976
                                        )

n_sim_points <- dim(sim_points)[1] # number simulated
n_est_detctions <- sum(weekly_counts$count) # number estimated by model
n_sim_points == n_est_detctions # same

head(sim_points)

# convert to SpatVector points
nws_points <- vect(sim_points, geom = c("x", "y"), crs = crs(spread_stk)) # projected 
nws_points <- project(nws_points, "EPSG:4326") # convert to lat/long


# back to dataframe, if needed
nws_point_latlong_df <- as.data.frame(nws_points, geom = "xy")
head(nws_point_latlong_df)
```


Save a copy of simulated point locations.  **Note:** This file is included in initial download from OSF above. 
```{r eval=FALSE, warning=FALSE, message=FALSE}
write_csv(nws_point_latlong_df, here("demo/sim_locations_2026-04-30.csv"))

```


#### Map Points  
Visual inspection of simulated locations for select weeks.    
```{r fig.width=15, fig.height=15, warning=FALSE, message=15}
quick_point_map(target_year=2022, target_week=10, 
                boundary_polys=states_proj, 
                point_data=sim_points)
```

```{r fig.width=15, fig.height=15, warning=FALSE, message=15}
quick_point_map(target_year=2024, target_week=30, 
                boundary_polys=states_proj, 
                point_data=sim_points)
```

```{r fig.width=15, fig.height=15, warning=FALSE, message=15}
quick_point_map(target_year=2026, target_week=2, 
                boundary_polys=states_proj, 
                point_data=sim_points)
```










