---
title: "Mesh"
description: "Mesh construction, tesselation, and defining the spatial domain"
format:
  html: 
    df-print: kable
    code-fold: show
    code-summary: "Hide code"
    code-overflow: wrap
    toc-title: Page Contents
    toc: true
    toc-depth: 2
    toc-location: right
    number-sections: false
    html-math-method: katex
    smooth-scroll: true
editor: source
editor_options: 
  chunk_output_type: console
---
```{=html}
<style type="text/css">

body, td {
   font-size: 11pt;
}
code.r{
  font-size: 9pt;
}
pre {
  font-size: 11pt
}
</style>
```


## Overview
The script demonstrates how to download model results archived on the Open Science Framework (OSF) and simulate New World screwworm (NWS) case detections.   
    
The downloaded data include`s spatial GeoTiff files depicting niche capacity and estimated spread for each week between January 2022 and February 2026, spatial boundaries for analysis and visualization, and a copy of previously simulated NWS cases.   

## Libraries  
Load needed libraries and packages.  
```{r, warning=FALSE, message=FALSE}
library(tidyverse) # data wrangling
options(dplyr.summarise.inform = FALSE)
library(here) # directory managment
library(pals) # color palettes
library(osfr) # interact with Open Science Framework (OSF, data repository)
library(terra) # spatial data manipulation
library(sf) # spatial data manipulation
library(spatstat) # spatial data manipulation
library(raster) # spatial data manipulation
select <- dplyr::select # deconflict dplyr w/raster  

# see https://www.r-inla.org/home
# install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
library(INLA) # mesh construction, inference
```

## Custom Functions 
Load customized functions.  
```{r, warning=FALSE, message=FALSE}
source(here("R/utilities.R"))
source_dir(here("R"))
```

## Download OSF Data  
Spatial GeoTIFF files (.tif) depicting NWS estimated niche capacity and spread are available on the [Open Science Framework](https://osf.io/uvqmw/).  Here we link the location and download the files to a demonstration directory (`/demo`).   
**Downloaded Files:**    
- *spread_GeoTIFF_2026-03-24.zip*: Zipped folder containing estimated NWS spread.       
- *niche_GeoTIFF_2026-03-24.zip*: Zipped folder containing niche-based estimates for NWS detections.   
- *study_polygons.gpkg*: Geopackage containing *sf* class multipolygons delineating political boundaries.    
- *sim_locations_2026-03-24.csv*: Simulated NWS detections; the same file is reproduced in this script.   
   

```{r, warning=FALSE, message=FALSE}
# file location on OSF
osf_project_location <- osf_retrieve_node("https://osf.io/uvqmw/")

# Project anem on OSF
osf_project_location$name 

# create demo directory
if (!dir.exists("demo")) dir.create("demo", recursive = TRUE)

# get data file names
osf_id <- osf_project_location %>%
  osf_ls_files()

# download all data
osf_download(osf_id,
             path = here("demo"),
             conflicts = "overwrite")
```


## Spatial Polygons
Read polygons defining jurisdictional boundaries
```{r, warning=FALSE, message=FALSE}
states_proj <- st_read(here("demo/study_polygons.gpkg"))
```

Simplify geometry to help define spatial extent of study area.  
```{r}
states_union <- st_union(states_proj) # dissolve states

# explode islands
poly_single <- st_cast(states_union, "POLYGON", warn = FALSE)

# calculate areas
areas <- st_area(poly_single)

# set threshold to remove smaller islands
threshold <- units::set_units(500, km^2)

# remove islands
states_simple <- poly_single[areas > threshold, ]
```


## Simulated NWS Cases
These are simulated NWS cases as simulated in the [Simulation tab]() of this site.
```{r}
sim_data_df <- read_csv(here("demo/sim_locations_2026-03-24.csv"))

nws_points <- vect(sim_data_df, geom = c("x", "y"), crs = "EPSG:4326") # raw are unprojected
nws_points <- project(nws_points, crs(states_proj)) # project to aea, units km
```


## Build Spatial Mesh
Define spatial domain through construction of a spatial triangulation (mesh). Requires use of the [**r-INLA** package](https://www.r-inla.org/home).  
```{r warning=FALSE, eval=TRUE}
# process requires classic spatial objects
points_sp <- as(nws_points, "Spatial")

# slight buffer distance to smooth edges
states_simple_buf <- st_buffer(states_simple, dist = 20)
simple_union <- st_union(states_simple_buf) # dissolve states

# convert to Spatial
dom_bnds <-  as(simple_union, "Spatial")

# to INLA format
dom_seg <- inla.sp2segment(dom_bnds)

set.seed(1976)
mesh.dom <- inla.mesh.2d(boundary = dom_seg, 
                        loc = points_sp,
                        cutoff = 25, 
                        max.edge = c(50, 200),
                        offset = c(100,250),
                        min.angle = 30) 

mesh.dom$n # node number
```

### View Mesh
Simulated NWS cases overlain as points. 
```{r fig.height=15, fig.width=15}
plot_mesh(mesh.dom)
```

## Natural Neighborhoods  
Apply tessellation to mesh vertices to identify area of support, or natural neighborhoods around each.  
```{r message=FALSE, warning=FALSE}
# mesh areas to polygons
mesh_polys <- convert_mesh_poly(mesh.dom)

# to spatvector and project
mesh_polys <- vect(mesh_polys)
crs(mesh_polys) <- crs(states_proj)

# calculate area
mesh_polys$area <- expanse(mesh_polys, unit = "km")
```

### Plot Neighborhoods  
Visual inspection of study area with neighborhoods color coded by geographic area and overlain with mesh and nodes.  
```{r warning=FALSE, message=FALSE, fig.width=15, fig.height=15}
mesh_polys_sf <- sf::st_as_sf(mesh_polys)

plot_duel_mesh(mesh_polys_sf, mesh.dom,
               bbox = c(xmin = -2500, # zoom to location extent
                        ymin = -2500, 
                        xmax = 1500, 
                        ymax = 1500))
```

Zoom in closer and remove ocean neighborhoods
```{r warning=FALSE, message=FALSE, fig.width=15, fig.height=15}
states_simple_sf <- st_as_sf(states_simple)
mesh_polys_sf_crop <- st_intersection(mesh_polys_sf, states_simple_sf)

plot_duel_mesh(mesh_polys_sf_crop, mesh.dom,
               bbox = c(xmin = -1500, # zoom location
                        ymin = -1100, 
                        xmax = -400, 
                        ymax = -50))
```


