---
title: "Tessellation"
subtitle: "Mesh Construction and Tessellation"
date: last-modified
format:
  html: 
    df-print: kable
    code-fold: show
    code-summary: "Show R Code"
    code-overflow: wrap
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    html-math-method: katex
    theme: cosmo
---
```{=html}
<style type="text/css">

body, td {
   font-size: 11pt;
}
code.r{
  font-size: 9pt;
}
pre {
  font-size: 11pt
}
</style>
```


## Overview
The script demonstrates spatial discretization of the study area as detailed in the associated manuscript. First, a triangulated mesh is constructed using administrative boundaries and simulated case point locations. Next, *natural neighborhoods* are defined using Voronoi tessellation around mesh vertices, such that each cell encapsulates the area each vertex. The resulting polygons define the spatial support and associated integration weights for numerical approximation of the spatial point-process model.  
  
::: {.callout-tip icon=false} 
#### Note
This script assumes that supporting data has already been downloaded to a directory called `/demo` and that NWS case detections have been simulated as described on the [Simulation](https://geoepi.github.io/hominivorax-geostat/simulation.html) page of this site. 
:::


---


## Libraries  
Load needed libraries and packages.  
```{r, warning=FALSE, message=FALSE}
library(tidyverse) # data wrangling
options(dplyr.summarise.inform = FALSE)
library(here) # directory managment
library(pals) # color palettes
library(osfr) # interact with Open Science Framework (OSF, data repository)
library(terra) # spatial data manipulation
library(sf) # spatial data manipulation
library(spatstat) # spatial data manipulation
library(raster) # spatial data manipulation
select <- dplyr::select # deconflict dplyr w/raster  

# see https://www.r-inla.org/home
# install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
library(INLA) # mesh construction, inference
```

## Custom Functions 
Load customized functions.  
```{r, warning=FALSE, message=FALSE}
source(here("R/load_preprocessing.R"))
load_preprocessing(here())
source(here("R/utilities.R"))
source(here("R/plot_mesh.R"))
source(here("R/plot_duel_mesh.R"))
```


## Spatial Polygons
Read polygons defining jurisdictional boundaries
```{r, warning=FALSE, message=FALSE}
states_proj <- st_read(here("demo/study_polygons.gpkg"))
```

Simplify geometry to help define spatial extent of study area.  
```{r warning=FALSE, message=FALSE}
states_union <- st_union(states_proj) # dissolve states

# explode islands
poly_single <- st_cast(states_union, "POLYGON", warn = FALSE)

# calculate areas
areas <- st_area(poly_single)

# set threshold to remove smaller islands
threshold <- units::set_units(500, km^2)

# remove islands
states_simple <- poly_single[areas > threshold]
```


## Simulated NWS Cases
These are NWS cases as simulated in the [Simulation tab](https://geoepi.github.io/hominivorax-geostat/simulation.html) of this site.
```{r warning=FALSE, message=FALSE}
sim_data_df <- read_csv(here("demo/sim_locations_demo.csv"))

nws_points <- vect(sim_data_df, geom = c("x", "y"), crs = "EPSG:4326") # raw are unprojected
nws_points <- project(nws_points, crs(states_proj)) # project to aea, units km
```


## Build Spatial Mesh
Define spatial domain through construction of a spatial triangulation (mesh). Requires use of the [**r-INLA** package](https://www.r-inla.org/home).  
   
Note that because the simulated points used in this script differ from the data presented in the associated paper, results here may be slightly different.     
```{r warning=FALSE, eval=TRUE, message=FALSE}
cfg_preprocess <- list(
  project = list(seed = 1976),
  study = list(projected_crs = sf::st_crs(states_proj)),
  mesh = list(boundary_buffer_km = 20, cutoff_km = 25, max_edge_km = c(50, 200),
              offset_km = c(100, 250), minimum_angle_degrees = 30,
              island_area_threshold_km2 = 500)
)
support <- build_spatial_support(states_proj, as.data.frame(nws_points, geom = "xy"), cfg_preprocess)
mesh.dom <- support$mesh
mesh.dom$n
```

### View Mesh
Simulated NWS cases overlain as points. 
```{r fig.height=15, fig.width=15, warning=FALSE, message=FALSE}
#| label: fig-mesh
#| fig-cap: "Triangulated mesh for the study area with simulates NWS case detection overlain."
#| 
plot_mesh(mesh.dom)
```

## Natural Neighborhoods  
Apply tessellation to mesh vertices to identify area of support, or natural neighborhoods around each. Here, this is done using the custom *convert_mesh_poly()* function.   
```{r message=FALSE, warning=FALSE, warning=FALSE, message=FALSE}
# mesh areas to polygons
mesh_polys <- terra::vect(support$mesh_polygons)
mesh_polys$area <- support$mesh_polygons$area_km2
```

### Plot Neighborhoods  
Visual inspection of study area with neighborhoods color coded by geographic area and overlain with mesh and nodes.  
```{r warning=FALSE, message=FALSE, fig.width=15, fig.height=15}
#| label: fig-tess1
#| fig-cap: "Tessellation around mesh nodes to identify natural neighborhoods.  Zoomed view shows major terrestrial boundaries, mesh, and simulated NWS case detections overlain on neighborhoods.  Neighborhoods are colored by geographic area."


mesh_polys_sf <- sf::st_as_sf(mesh_polys)

plot_duel_mesh(mesh_polys_sf, mesh.dom,
               bbox = c(xmin = -2500, # zoom to location extent
                        ymin = -2500, 
                        xmax = 1500, 
                        ymax = 1500))
```

Zoom in closer and remove ocean neighborhoods to improve visualization.  
```{r warning=FALSE, message=FALSE, fig.width=15, fig.height=15}
#| label: fig-tess2
#| fig-cap: "Tessellation around mesh nodes to identify natural neighborhoods.  Zoomed view shows mesh and simulated NWS case detections overlain on neighborhoods.  Neighborhoods are colored by geographic area, with those over major waterbodies have been removed for clarity."
#| 
states_simple_sf <- st_as_sf(states_simple)
mesh_polys_sf_crop <- st_intersection(mesh_polys_sf, states_simple_sf)

plot_duel_mesh(mesh_polys_sf_crop, mesh.dom,
               bbox = c(xmin = -1500, # zoom location
                        ymin = -1100, 
                        xmax = -400, 
                        ymax = -50))
```

### Save Mesh and Tesselation  
Save the triangulated mesh and areal tessellation for the next step. 
```{r warning=FALSE, message=FALSE, fig.width=15, fig.height=15}
# save mesh
saveRDS(mesh.dom, here("demo/mesh.rds"))

# copy of natural neighborhoods
writeVector(mesh_polys, here("demo/neighboorhoods.gpkg"), overwrite = TRUE)

```

::: {.callout-tip icon=false} 
#### Next Step
Once the triangulated mesh and natural neighborhoods have been created and saved, proceed to proceed to the [Response](https://geoepi.github.io/hominivorax-geostat/response.html) page of this site. 
:::

